### 2025-02-24 paper/sections/other/counterfactuals.qmd

>>>>

Gradient-based counterfactual search was originally proposed by @wachter2017counterfactual. It generally solves the following unconstrained objective,

$$
\begin{aligned}
\min_{\mathbf{z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(\mathbf{M}_{\theta}(g(\mathbf{z}^\prime)),\mathbf{y}^+)}+ \lambda {\text{cost}(g(\mathbf{z}^\prime)) }  \right\} 
\end{aligned} 
$$

where $g: \mathcal{Z} \mapsto \mathcal{X}$ is an invertible function that maps from the $L$-dimensional counterfactual state space to the feature space and $\text{cost}(\cdot)$ denotes one or more penalties that are used to induce certain properties of the counterfactual outcome. As above, $\mathbf{y}^+$ denotes the target output and $\mathbf{M}_{\theta}(\mathbf{x})$ returns the logit predictions of the underlying classifier for $\mathbf{x}=g(\mathbf{z})$.

For all generators used in this work we use standard logit crossentropy loss for $\text{yloss}(\cdot)$. All generators also penalize the distance ($\ell_1$-norm) of counterfactuals from their original factual state. For *Generic* and *ECCo*, we have $\mathcal{Z}:=\mathcal{X}$ and $g(\mathbf{z})=g(\mathbf{z})^{-1}=\mathbf{z}$, that is counterfactual are searched directly in the feature space. Conversely, *REVISE* traverses the latent space of a variational autoencoder (VAE) fitted to the training data, where $g(\cdot)$ corresponds to the decoder [@joshi2019realistic]. In addition to the distance penalty, *ECCo* uses an additional penalty component that regularizes the energy associated with the counterfactual, $\mathbf{x}^\prime$ [@altmeyer2024faithful].
